Supplementary Information: Tropical tree drought sensitivity is jointly shaped by drought characteristics, species adaptations, and individual microenvironments

### 2025-06-27 doc/manuscript_hkk_drought_sensitivity.Rmd
updated text with minor edits and word count

---
title: ''
output: 
    bookdown::word_document2: 
      #reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
      reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

@@

**Number of Words in Abstract:** 146  
**Number of Words in Main Text:** 5048  
**Number of References:** 55  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

**Corresponding author address:**  
Krishna Anujan,
ForestGEO Ecosystems and Climate Lab,  
Smithsonian's National Zoo and Conservation Biology Institute,  
Front Royal, VA 22630  
USA  
Phone: 
Email: anujank@si.edu

@@

<!---CONCLUSIONS---->
We show that, in a tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment (Table 1).
Analysing three droughts in a 14-year dendrometer band time series, we found lower-than-average growth in drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
Across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to water access (TWI, DBH) under differing drought conditions.
Despite lower extremes of temperature, vapor pressure deficit, and number of dry days (Fig. 2a), severe drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`). 
Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species showed greater growth declines in wetter sites in 2015 while evergreen species grew less in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b). 
Although trees with more exposed crowns were consistently more sensitive to drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a). 
It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines during wet-season droughts (Fig `r fig_dag`). 
<!-- we can talk about it ... but I don't think that we can know that larger trees have deeper roots.  -->
Taken together, these results indicate that combined effects of drought characteristics, species traits, and individual microenvironments produce important heterogeneity in individual and species drought responses of tropical trees.
Divergent effects of species and individuals to droughts have  may contribute to the maintenance of species diversity and ecosystem stability.

@@

<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses here compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different. 
We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to maximise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironments.
Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   -->

@@

<!-- contributions / disentangling exposure from DBH-->
<!-- One variable examined here -- crown exposure -- had a consistent effect across the two droughts (Table 1, Figs. `r fig_microenvs`, `r fig_dag`).  -->
Trees with more exposed crowns exhibited greater growth declines, while sub-canopy trees grew more than average during the 2010 drought.
This finding is consistent with the concept that larger trees tend to suffer more during drought [@bennett_larger_2015], but adds the important insight that it is crown position -- as opposed to size *per se* -- that is primarily responsible for this pattern.
The consistent negative effect of crown position aligns with pronounced gradients in drought stress across forest vertical profiles [@vinod_thermal_2023].
We show that residual effects of size, presumably including root water access and biophysical challenges linked to height itself, may act in different directions depending on the drought characteristics (Figs. `r fig_microenvs`, `r fig_dag`).
These findings imply that while canopy and emergent trees experience the greatest stress and the most negative growth responses, a suite of species traits and microenvironmental characteristics modify growth responses in interaction with drought characteristics.

<!----Contributions beyond---->
Despite consistent observations of plot-level growth declines across drought events, we show that this scale obscures a variety of species- and tree-level ecological responses that are heterogeneous across droughts. 
An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig. S5).
More generally, tree-ring records likely present a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.
<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. -->
Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies make particular species and trees resistant to some droughts but vulnerable to others. 
These distinct responses of species and individuals to droughts suggests potential spatiotemporal insurance effects among tropical tree response, in alignment with recent global analyses [@liu_drought_2022].

>>>>

---
title: ''
output: 
    bookdown::word_document2: 
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
      #reference_docx: word-styles-reference-singlespace.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
    #pdf_document:
    # engine: xelatex 
  #officedown::rdocx_document
bibliography: growth-precip-thailand.bib
# csl: new-phytologist.csl
csl: ecology-letters.csl 
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

@@

**Number of Words in Abstract:** 146  
**Number of Words in Main Text:** 5028  
**Number of References:** 55  
**Number of Figures:** 5  
**Number of Tables:** 1  
**Number of Text boxes:** 0

**Corresponding author address:**  
Krishna Anujan,
ForestGEO Ecosystems and Climate Lab,  
Smithsonian's National Zoo and Conservation Biology Institute,  
Front Royal, VA 22630  
USA  
Phone: +1 (202) 633-1836  
Email: anujank@si.edu

@@

<!---CONCLUSIONS---->
We show that, in a tropical seasonal forest, the drought sensitivity of tree growth is jointly and interactively shaped by drought characteristics, species adaptations, tree size and microenvironment (Table 1).
Analysing three droughts in a 14-year dendrometer band time series, we found lower-than-average growth during drought years, resulting in lower plot-level growth (Fig `r fig_drought_responses`). 
Across droughts, there was little consistency in species- or individual-level drought responses, reflecting dissimilar effects of a key species trait (deciduousness) and variables linked to tree drought experience (TWI, DBH, CII) under differing drought conditions.
Despite lower extremes of temperature, vapor pressure deficit, and number of dry days (Fig. 2a), severe drought in the wet season (2015) elicited stronger responses across trees, especially deciduous species, which are expected to have stronger seasonality in growth phenology (Fig `r fig_species`). 
Interestingly, in 2015 and 2020 (but not 2010), evergreen and deciduous species had divergent responses, especially along a wetness gradient; deciduous species showed greater growth declines in wetter sites in 2015 while evergreen species grew less in drier sites in 2020 (Fig `r fig_species`b, Fig `r fig_microenvs`b). 
Although trees with more exposed crowns tended to grow less under drought, additional effects of DBH -- presumably including rooting volume and depth -- were dissimilar between the drought events (Fig `r fig_microenvs`a). 
It is possible that water availability buffered exposure-associated growth declines through different mechanisms in the different droughts; larger trees fared better in the dry-season drought while trees in wetter sites suffered smaller growth declines during wet-season droughts (Fig `r fig_dag`). 
<!-- we can talk about it ... but I don't think that we can know that larger trees have deeper roots.  -->
Taken together, these results indicate that combined effects of drought characteristics, species traits, and individual microenvironments produce important heterogeneity in individual and species drought responses of tropical trees.
Divergent effects of species and individuals to droughts may contribute to the maintenance of species diversity and ecosystem stability.

@@

<!----Limitations in generalisation/ species examined here may have greater diversity of drought adaptaitons than elsewhere in the tropics---->
Periodic hot/dry conditions in tropical dry forests could create selective pressures for adaptative variation towards drought resistance potentially resulting in a greater diversity of drought responses here compared to other bioclimatic zones where adaptive pressures and thus ecological variability are different. 
We demonstrate that species identity and a key species adaptation (decidousness) interact with drought characteristics and microenvironment to shape drought responses in a dry seasonal evergreen forest.
Strong seasonality in tropical dry forests exerts selective pressures on hydraulic strategies resulting in diversification along the hydraulic safety-efficiency axis with diverse strategies to optimise tree vital rates [@gonzález-m._diverging_2021;@oliveira_linking_2021]. 
The degree of adaptive variation along the water-use axis might be comparable to other dry climatic systems (e.g. temperate drylands reported in @vasey_intraspecific_2022), but may be distinct from the wet/everwet tropics, where tree-environment relations and evolutionary strategies may be substantially different. 
Paralleling the rarity of long-term records of annual tree growth in less seasonal tropical forests [@zuidema_tropical_2022], we know exceedingly little about how tropical tree species with different traits respond to different types of drought, and how these responses may be modified by microenvironments.
Again, continued expansion of dendrometer band records and ecophysiological studies will be essential to understanding the diversity and complexity of tropical tree drought responses, as well as if and how this diversity may promote community-level insurance effects under drought across tropical forests in general.
<!--Therefore, we caution that our results suggestive of community-level insurance effects under drought be interpreted carefully across forest types that may be structured primarily by nutrient limitation, fires, herbivory or pathogens. 
However, our inference on the effects of drought timing could be generalisable across wet and dry tropics in the light of emerging evidence that despite low seasonality in climatic conditions, everwet forest species also show seasonal growth patterns [@giraldo_tree_2023].   -->

@@

<!-- contributions / disentangling exposure from DBH-->
<!-- One variable examined here -- crown exposure -- had a consistent effect across the two droughts (Table 1, Figs. `r fig_microenvs`, `r fig_dag`).  -->
Trees with more exposed crowns exhibited greater growth declines, while sub-canopy trees grew more than average during the 2010 drought.
This finding is consistent with observations that larger trees tend to suffer more during drought [@bennett_larger_2015], but adds the important insight that it is crown position -- as opposed to size *per se* -- that is primarily responsible for this pattern.
The consistent negative direction of the effect of crown position aligns with pronounced gradients in drought stress across forest vertical profiles [@vinod_thermal_2023].
We show that residual effects of size, presumably including root water access and biophysical challenges linked to height itself, may act in different directions depending on the drought characteristics (Figs. `r fig_microenvs`, `r fig_dag`).
These findings imply that while canopy and emergent trees experience the greatest stress and the most negative growth responses, a suite of species traits and microenvironmental characteristics modify growth responses in interaction with drought characteristics.

<!----Contributions beyond---->
Despite consistent observations of plot-level growth declines across drought events, we show that this scale obscures a variety of species- and tree-level ecological responses that are heterogeneous across droughts. 
An important implication is that drought responses inferred from tree-ring records derived from the few species known to form reliable, dateable annual rings (n=4 at HKK; @vlam_temperature_2014) are unlikely to be representative of the drought responses of the entire community.
For example, one of the few annual-ring-forming species at HKK, deciduous *Afzelia xylocarpa*, had the most distinct response across the three events; it's growth nearly doubled in the 2010 and 2020 droughts but was severely reduced in the wet-season drought (Fig. S5).
More generally, tree-ring records likely present a biased picture of drought responses, as ring formation is caused by dry-season dormancy (often associated with deciduous leaf habit) and sampling often targets large trees with exposed crowns and dry microenvironments [@speer_fundamentals_2010]-- all factors shown here to <!--directly and interactively -->influence drought sensitivity.
<!-- More generally, past analyses of tree ring and census estimates of drought sensitivity in seasonally dry tropical forests may have been limited towards fast growing (measurable using tape) or few dry-associated species (that form annual tree rings) - and therefore capturing only the more dramatic effects. -->
Our analyses reveal a complex heterogeneity of drought responses, including the contrasting influence of deciduousness in three different droughts, showing that ecological and evolutionary strategies can make particular species and trees resistant to some droughts but vulnerable to others. 
These distinct responses of species and individuals to droughts suggests potential spatiotemporal insurance effects among tropical tree responses, in alignment with recent global analyses [@liu_drought_2022].
